Transient Corotating Clumps Around Adolescent Low-Mass Stars: Four Years of Complex Quasiperiodic Variables from TESS

### 2023-08-18 paper/ms.tex
ms: fix typos

\author[0000-0002-0531-1073]{H. Isaacson}
\affiliation{\berkeley}

\author[0000-0001-8638-0320]{A. W. Howard}
\affiliation{\caltech}

@@

\begin{enumerate}[leftmargin=*]
	%
    \item CQVs are quasiperiodic.  The mean periods remain fixed over
      the $>$1{,}000-cycle baseline of available observations; but the
      light curve shapes always evolve (Figure~\ref{fig:evoln}).
  %
    \item The same CQV can show multiple periods simultaneously.  LP
      12-502, for instance, showed dips with four distinct periods
      within $\pm 0.3\%$ of its fundamental period, sometimes
      simultaneously, and each lasting for up to 50 cycles
      (Figure~\ref{fig:lpriver0}).
  %
    \item CQVs evolve over timescales that are both secular
      ($>$100\,cycles) and impulsive ($<$1 cycle).  Dip growth seems
      to happen over durations of at least ten cycles, and slow dip
      decay can also occur.  ``State-switches'' correspond to dips to
      collapsing instantaneously, and are almost always linked with
      observed optical flares.  This suggests that the occulting
      material is sensitive to sudden changes in the magnetic field.
  %
    \item The rate of dip evolution can be used to place a
      model-dependent constraint on how much material is either being
      accreted or ejected during the state changes
      (Section~\ref{subsec:massflux}).  Order of magnitude estimates
      require at least an asteroid belt's worth of dust
      ($10^{-4}$\,$M_\oplus$) over $10^8$ years, or
      $\approx$$10^{-2}$\,$M_\oplus$ if the occulting material is gas.
  %
    \item The CQV phenomenon persists for $\gtrsim$150\,Myr, based on
      the existence of multiple CQVs in AB Dor, the Pleiades, and
      Psc-Eri (Section~\ref{subsec:starprops}).  It may even extend to
      200\,Myr, based on the one CQV we found in the Carina Near
      moving group (TIC~294328887; $\approx$200\,Myr).  However the
      lack of detected CQVs in the Hyades and Praesepe suggests that
      the lifetime of the phenomenon is limited to the first few
      hundred million years.
	%
    \item The duty cycle for CQVs seems to be $\approx$$75$\%, based
      on the fraction of bona fide CQVs that either turned on or turned
      off during TESS re-observations, two years after the initial
      observation.
  %
    \item Most CQVs are M dwarfs with masses
      0.1-0.5\,$M_\odot$.  Two sources, TIC~405754448 and
      TIC~405910546, have masses that appear to exceed the M dwarf
      limit.  However both are potentially binaries, and this may
      confuse our ability to accurate identify the source of the CQV
      signal (Section~\ref{subsec:massive}).  We encourage additional
      scrutiny of these objects in future work.
  %
    \item Surprising analogs to the CQVs exist in magnetic B stars,
      but only those with multipolar field topologies.  Since most
      magnetic B stars have dipolar magnetic fields, this suggests
      that the CQV dips and warps are similarly being sculpted by the
      stellar magnetic fields, and that the magnetic fields themselves
      are probably also multipolar.
  %
    \item The closest CQVs to the Sun are at distances of 15-20\,pc,
      and the brightest have $V$$\approx$12 ($J$$\approx$7.5).  We
      have found most of them in this work, since our CQV sample was
      $\gtrsim$80\% complete within 30\,pc.  The lack of CQVs in the
      volume-complete $<$15\,pc sample of 0.1-0.3\,$M_\odot$ stars
      analyzed by \citet{2021AJ....161...63W} is consistent with this
      estimate.  Expanding our analysis of the TESS data to the full
      frame images would yield a truly volume-limited selection
      function, and would expand the CQV census by about a factor of
      two within 50\,pc, and by a factor of ten within 100\,pc.
\end{enumerate}

@@

In both the gas and dust scenarios, CQVs are preferentially viewed
edge-on.  This implies that after correcting for the line-of-sight
inclination, roughly one third of low mass stars \citep[those that
rotate rapidly enough][]{2022AJ....163..144G} could trap circumstellar
material in the same way.  It also suggests that CQVs may
preferentially show transiting planets at larger distances than the
corotating material, though this may be dependent on whether the
magnetic and stellar spin axes tend to be aligned.  Given these
points, observational follow-up work should include searching for
outer transiting planets, and measuring equatorial velocities in order
to test whether the stellar inclination angles are indeed
preferentially edge-on.  Any source of empirical information on the
stellar magnetic field, whether from the Zeeman effect
\citep[e.g.][]{2021A&ARv..29....1K} or perhaps radio emission
\citep[e.g.][]{2015Natur.523..568H}, could also help clarify the
strength of the magnetospheres for these objects.

@@

%{\bf AUTHOR CONTRIBUTION STATEMENTS: PLEASE REVISE IF APPROPRIATE)}
LGB conceived the project and performed the dip-counting search, light curve classification, cluster membership, SED, variability, and secondary-period analyses.
RJ and SR performed the Fourier-based analysis and contributed to light curve classification.
LR cross-examined the light curve classification, and contributed an independent SED analysis.
AD identified the magnetic B star connection.
LAH contributed to project design and to the interpretation of the light curves.
G\'AB acquired and maintained the servers used to run the dip-finding pipeline.
-----
{\bf JNW, MNG, GRR, HI, and AH to confirm coauthorship \&/or to help out on interpretation!}
HI and AH designed and maintain the architecture used to acquire and process the HIRES spectra.
All authors assisted in manuscript writing and revision.

@@

Figures~\ref{fig:tic3006timegroups} and~\ref{fig:tic3006river} show
120-second cadence data for TIC~300651846, a CQV in the TESS
continuous viewing zone.  If it were not for the existence of
TIC~402980664, this source would probably have received greater
attention.  With the exception of a few sectors, TESS data will exist
for TIC~300651846 for at least Sectors 1-12, 27-39, and 61-69.  While
most of the available data will exist in the full frame images, in
Figures~\ref{fig:tic3006timegroups} and~\ref{fig:tic3006river} focus
only on the currently available 120-second cadence data.

@@

TESS was the first instrument to show that CQV light curves contain
power at timescales of a few minutes
\citep{2019A&A...625L..13Z,2022AJ....163..144G}.  This advance was
enabled by the fifteen-fold faster cadence in the TESS 2-minute data,
relative to K2.  A logical follow-up is to ask whether the periodic
components of the CQV light curves contain power at timescales below
one minute.  Between 2020 and 2021, we observed 10 CQVs at 20-second
cadence with TESS in order to explore this question (TESS DDT029).
The stars were TICs 142173958, 146539195, 24518895, 276453848,
264599508, 363963079, 144486786, 408188366, 300651846, 262400835.
These sources were selected from CQVs known at the time to have short
periods and sharp features when observed at 2-minute cadence.
Comparing the 20-second to 120-second data for these stars (data
available on MAST), we concluded that these CQVs did not contain
appreciable power at timescales shorter than a few minutes.

@@

We acquired iodine-free reconnaissance spectra using Keck/HIRES for
three CQVs.  The goals were to determine the chromospheric activity
levels, and to check for indications of either accretion or
spectroscopic binarity.  We acquired a 15 minute exposure of
TIC~146539195 on 2023 January 3, a 15 minute exposure of TIC~264599508
on 2023 January 9, and a 30 minute exposure of TIC~402980664 on 2023
July 10.  The acquisition and analysis followed the usual techniques
of the California Planet Survey \citep{2010ApJ...721.1467H}.
Figure~\ref{fig:speccutouts} shows cutouts from the resulting spectra,
centered on the \ion{Ca}{2} HK windows, H$\alpha$, and the \ion{Li}{1}
6708\,\AA\ doublet.  The \ion{Ca}{2} H emission line is blended with
H$\epsilon$.  While a more detailed analysis will be left for future
work, these spectra confirm previous understanding established by
\citet{2017AJ....153..152S} that the stars are chromospherically
active M dwarfs in the ``weak-lined'' T Tauri regime
\citep[e.g.][Figure~15]{2019AJ....157...85B}.  Their H$\alpha$
equivalent widths, at $\approx$14\,\AA, $\approx$3\AA, and
$\approx$8\,\AA (for TIC~264599508, 146539195, and 402980664
respectively) are consistent with purely chromospheric emission.  The
blue excess in TIC~264599508 could be explained by a second unresolved
star; the TESS light curve for this source shows both the 7.90\,hr CQV
signal, and a 9.00\,hr rotation signal with comparable amplitude.

>>>>

% PENDING DISCUSSION!
%\author[0000-0002-0531-1073]{H. Isaacson}
%\affiliation{\berkeley}
%
%\author[0000-0001-8638-0320]{A. W. Howard}
%\affiliation{\caltech}

@@

\begin{enumerate}[leftmargin=*]
	%
    \item CQVs are quasiperiodic.  The mean periods remain fixed over
      the $>$1{,}000-cycle baseline of available observations; but the
      light curve shapes always evolve (Figure~\ref{fig:evoln}).
  %
    \item The same CQV can show multiple periods simultaneously.  LP
      12-502, for instance, showed dips with four distinct periods
      within $\pm 0.3\%$ of its fundamental period, sometimes
      simultaneously, and each lasting for up to 50 cycles
      (Figure~\ref{fig:lpriver0}).
  %
    \item CQVs evolve over timescales that are both secular
      ($>$100\,cycles) and impulsive ($<$1 cycle).  Dip growth seems
      to happen over durations of at least ten cycles, and slow dip
      decay can also occur.  ``State-switches'' correspond to dips to
      collapsing instantaneously, and are almost always linked with
      observed optical flares.  This suggests that the occulting
      material is sensitive to sudden changes in the magnetic field.
  %
    \item The rate of dip evolution can be used to place a
      model-dependent constraint on how much material is either being
      accreted or ejected during the state changes
      (Section~\ref{subsec:massflux}).  Order of magnitude estimates
      require at least an asteroid belt's worth of dust
      ($10^{-4}$\,$M_\oplus$) over $10^8$ years, or
      $\approx$$10^{-2}$\,$M_\oplus$ if the occulting material is gas.
  %
    \item The CQV phenomenon persists for $\gtrsim$150\,Myr, based on
      the existence of multiple CQVs in AB Dor, the Pleiades, and
      Psc-Eri (Section~\ref{subsec:starprops}).  It may even extend to
      200\,Myr, based on the one CQV we found in the Carina Near
      moving group (TIC~294328887; $\approx$200\,Myr).  However the
      lack of detected CQVs in the Hyades and Praesepe suggests that
      the lifetime of the phenomenon is limited to the first few
      hundred million years.
	%
    \item The duty cycle for CQVs seems to be $\approx$$75$\%, based
      on the fraction of bona fide CQVs that either turned on or turned
      off during TESS re-observations, two years after the initial
      observation (Figure~\ref{fig:evoln}).
  %
    \item Most CQVs are M dwarfs with masses
      0.1-0.5\,$M_\odot$.  Two sources, TIC~405754448 and
      TIC~405910546, have masses that appear to exceed the M dwarf
      limit.  However both are potentially binaries, and this may
      confuse our ability to accurate identify the source of the CQV
      signal (Section~\ref{subsec:massive}).  We encourage additional
      scrutiny of these objects in future work.
  %
    \item Surprising analogs to the CQVs exist in magnetic B stars,
      but only those with multipolar field topologies.  Since most
      magnetic B stars have dipolar magnetic fields, this suggests
      that the CQV dips and warps are similarly being sculpted by the
      stellar magnetic fields, and that the magnetic fields themselves
      are probably also multipolar.
  %
    \item The closest CQVs to the Sun are at distances of 15-20\,pc,
      and the brightest have $V$$\approx$12 ($J$$\approx$7.5).  We
      have found most of them in this work, since our CQV sample was
      $\gtrsim$80\% complete within 30\,pc.  The lack of CQVs in the
      volume-complete $<$15\,pc sample of 0.1-0.3\,$M_\odot$ stars
      analyzed by \citet{2021AJ....161...63W} is consistent with this
      estimate.  Expanding our analysis of the TESS data to the full
      frame images would yield a truly volume-limited selection
      function, and would expand the CQV census by about a factor of
      two within 50\,pc, and by a factor of ten within 100\,pc.
\end{enumerate}

@@

In both the gas and dust scenarios, CQVs are preferentially viewed
edge-on.  This implies that after correcting for the line-of-sight
inclination, roughly one third of low mass stars \citep[those that
rotate rapidly enough;][]{2022AJ....163..144G} could trap circumstellar
material in the same way.  It also suggests that CQVs may
preferentially show transiting planets at larger distances than the
corotating material, though this may be dependent on whether the
magnetic and stellar spin axes tend to be aligned.  Given these
points, observational follow-up work should include searching for
outer transiting planets, and measuring equatorial velocities in order
to test whether the stellar inclination angles are indeed
preferentially edge-on.  Any source of empirical information on the
stellar magnetic field, whether from the Zeeman effect
\citep[e.g.][]{2021A&ARv..29....1K} or perhaps radio emission
\citep[e.g.][]{2015Natur.523..568H}, could also help clarify the
strength of the magnetospheres for these objects.

@@

%{\bf AUTHOR CONTRIBUTION STATEMENTS: PLEASE REVISE IF APPROPRIATE)}
LGB conceived the project and performed the dip-counting search, light curve classification, cluster membership, SED, variability, and secondary-period analyses.
RJ and SR performed the Fourier-based analysis and contributed to light curve classification.
LR cross-examined the light curve classification, and contributed an independent SED analysis.
AD identified the magnetic B star connection.
LAH contributed to project design and to the interpretation of the light curves.
G\'AB acquired and maintained the servers used to run the dip-finding pipeline.
-----
{\bf JNW, MNG, GRR to confirm coauthorship \&/or to help out on interpretation!}
% Pending!
%HI and AH designed and maintain the architecture used to acquire and process the HIRES spectra.
All authors assisted in manuscript writing and revision.

@@

Figures~\ref{fig:tic3006timegroups} and~\ref{fig:tic3006river} show
120-second cadence data for TIC~300651846, a CQV in the TESS
continuous viewing zone.  If it were not for the existence of
TIC~402980664, this source would probably have received greater
attention.  With the exception of a few sectors, TESS data will exist
for TIC~300651846 for at least Sectors 1-12, 27-39, and 61-69.  While
most of the available data exist in the full frame images,
Figures~\ref{fig:tic3006timegroups} and~\ref{fig:tic3006river} focus
only on the currently available 120-second cadence data.

@@

TESS was the first instrument to show that CQV light curves contain
power at timescales of a few minutes
\citep{2019ApJ...876..127Z,2022AJ....163..144G}.  This advance was
enabled by the fifteen-fold faster cadence in the TESS 2-minute data,
relative to K2.  A logical follow-up is to ask whether the periodic
components of the CQV light curves contain power at timescales below
one minute.  Between 2020 and 2021, we observed 10 CQVs at 20-second
cadence with TESS in order to explore this question (TESS DDT029).
The stars were TICs 142173958, 146539195, 24518895, 276453848,
264599508, 363963079, 144486786, 408188366, 300651846, 262400835.
These sources were selected from CQVs known at the time to have short
periods and sharp features when observed at 2-minute cadence.
Comparing the 20-second to 120-second data for these stars (data
available on MAST), we concluded that these CQVs did not contain
appreciable power at timescales shorter than a few minutes.

@@

We acquired iodine-free reconnaissance spectra using Keck/HIRES for
three CQVs.  The goals were to determine the chromospheric activity
levels, and to check for indications of either accretion or
spectroscopic binarity.  We acquired a 15 minute exposure of
TIC~146539195 on 2023 January 3, a 15 minute exposure of TIC~264599508
on 2023 January 9, and a 30 minute exposure of TIC~402980664 on 2023
July 10.  The acquisition and analysis followed the usual techniques
of the California Planet Survey \citep{2010ApJ...721.1467H}.
Figure~\ref{fig:speccutouts} shows cutouts from the resulting spectra,
centered on the \ion{Ca}{2} HK windows, H$\alpha$, and the \ion{Li}{1}
6708\,\AA\ doublet.  The \ion{Ca}{2} H emission line is blended with
H$\epsilon$.  While a more detailed analysis will be left for future
work, these spectra confirm previous understanding established by
\citet{2017AJ....153..152S} that the stars are chromospherically
active M dwarfs in the ``weak-lined'' T Tauri regime
\citep[e.g.][Figure~15]{2019AJ....157...85B}.  Their H$\alpha$
equivalent widths, at $\approx$14\,\AA, $\approx$3\AA, and
$\approx$8\,\AA\ (for TIC~264599508, 146539195, and 402980664
respectively) are consistent with purely chromospheric emission.  The
blue excess in TIC~264599508 could be explained by a second unresolved
star; the TESS light curve for this source shows both the 7.90\,hr CQV
signal, and a 9.00\,hr rotation signal with comparable amplitude.
